---
title: "Assignment 1: Group 45"
author: "Daniel Engbert, Rik Timmer, Koen van der Pool"
date: "11 February 2023"
output: pdf_document
fontsize: 11pt
highlight: tango
editor_options: 
  markdown: 
    wrap: 72
---

```{r, echo=FALSE}
#options(digits=3)

# helper function to check normality throughout this assignment
checkNorm = function(data, name, unit="") {
  xlab = name
  if (unit != "") {
    xlab = sprintf("%s (%s)", name, unit)
  }
  hist(data, main=sprintf("Histogram of %s", name), xlab=xlab)
  qqnorm(data, main=sprintf("Normal qqplot of %s", name))
  res = shapiro.test(data)
  print(sprintf("Shapiro-Wilk normality p-value for %s: %.3f", name, res$p.value))
  return(res$p.value)
}

printPval = function(pval) {
  print(sprintf("p-value = %.3f", pval))
}
```

Note: we made a function `checkNorm()` which prints a histogram, qqplot,
and p-value from the shapiro-wilk normality test. And we made a function
`printPval()` which simply prints a given p-value to 3 significant
figures. We utilize both functions throughout this assignment.

## Exercise 1: Trees

### 1 a)

```{r}
trees = read.table("treeVolume.txt", header=T)
trees
```

## Exercise 2: Expenditure on criminal activities

### 1 a)

```{r}
crimes = read.table("expensescrime.txt", header=T)
crimes
```

## Exercise 3: Titanic

### 1 a)

```{r}
titanic = read.table("titanic.txt", header=T)
titanic
```

## Exercise 4: Military Coups

### 1 a)

```{r}
coups = read.table("coups.txt", header=T)
coups
```
