---
title: "Assignment template"
author: "Group 45: Daniel Engbert, Rik Timmer, Koen van der Pool"
date: "11 February 2023"
output: pdf_document
fontsize: 11pt
highlight: tango
editor_options: 
  markdown: 
    wrap: 72
---

## Exercise 1: Birthweights

<!--# See template_assign_edda.Rmd for full tips and info for formatting this document -->

```{r}
data = read.table("birthweight.txt", header=T)
par(mfrow=c(1,2))
data
# checking normality
hist(data$birthweight, freq=F)
qqnorm(data$birthweight)
shapiro.test(data$birthweight)

res = t.test(data$birthweight, conf.level=0.96)
res
sd(data$birthweight)

tval = qt(0.98, length(data$birthweight))
m = 50
n = (tval * sd(data$birthweight) / m) ** 2
n
#sprintf("96% confidence interval: [%.4f, %.4f]", res$conf.int[1], res$conf.int[2])

```

**a)**

The birthweight data appears to be normal based on a normal-appearing
histogram, the straight line in the qqplot, and the shapiro-wilk
normality test (having \$p=0.8995 \> 0.05\$).

$$
m = \frac{t*s}{\sqrt{n}} \rightarrow n = (\frac{t * s}{m})^2
$$

The equation above, gives the sample size, \$n\$, needed for the
confidence interval to have a width of 100 (meaning \$m=50\$), where $t$
is the t-score for the quantile 0.02 (such that both tails of the
distribution have total area 0.04). Our calculations compute \$n =
832.32\$, which rounded indicates the min sample size is $n=833$ (for a
96% CI).

TODO: see slide 9 for bootstrap CI code

## Exercise 2: Cholesterol

```{r,echo=FALSE,fig.height=3.5}
data = read.table("cholesterol.txt", header=T)
data
```

## Exercise 3: Diet

```{r}
data = read.table("diet.txt", header=T)
data
```

## Exercise 4: Yield of Peas

```{r}
library(MASS)
data = npk
data
```
